---
title: "Correlating residuals with interaction information"
author: "Elena Quintero"
# format: gfm
format: html  
self-contained: true
---

Load data

```{r}
#| message: false
#| warning: false
library(here)
library(tidyverse)
library(ggplot2)
library(ggdist)
```

```{r}
#| message: false
load(here::here("hmsc/output/cooccurrence_matrix_non_spat.RData"))

ints <- read_csv(here::here("interaction/data/Swiss_web_gen_strict.csv"))
metaweb <- read_csv(here::here("interaction/data/Swiss_metaweb_gen_relaxed.csv"))

taxa <- read_csv(here::here("interaction/taxa_list_hmsc.csv"), col_names = FALSE)
butterflies <- taxa$X1[!is.na(taxa$X1)] |> unique()
plants <- taxa$X3[!is.na(taxa$X3)] |> unique() 
```

## Correlation JSDM residuals and species interactions

```{r}
residuals_mat <- OmegaCor[[1]]$mean

# keep only upper triangle of the matrix
residuals_mat[lower.tri(residuals_mat, diag = TRUE)] <- NA

residuals_long <- residuals_mat |>
    as.data.frame() |>
    rownames_to_column("bf_taxa") |>
    pivot_longer(cols = -bf_taxa, names_to = "pl_taxa", values_to = "residuals") |> 
    filter(!is.na(residuals)) |> 
    filter(bf_taxa %in% butterflies & pl_taxa %in% plants)

# checking dimenstion
nrow(residuals_long)
163*165 #plant taxa times animal taxa
```

Load interaction data and metaweb data, and join with residuals.
```{r}
ints_long <- ints |> 
    rename(pl_taxa = ...1) |>
    pivot_longer(cols = -pl_taxa, names_to = "bf_taxa", values_to = "interaction") |> 
    #Change to "Polygonia_c.album" "Satyrium_w.album" 
    mutate(bf_taxa = str_replace(bf_taxa, "-", "."))
nrow(ints_long) # more interactions than in residuals df, because some species are not in the model
# setdiff(unique(ints_long$bf_taxa), unique(residuals_long$bf_taxa))

metaweb_long <- metaweb |> 
    rename(pl_taxa = ...1) |>
    pivot_longer(cols = -pl_taxa, names_to = "bf_taxa", values_to = "int_potential") |> 
    #Change to "Polygonia_c.album" "Satyrium_w.album" 
    mutate(bf_taxa = str_replace(bf_taxa, "-", "."))
nrow(metaweb_long)

df <- residuals_long |> 
    left_join(ints_long, by = c("bf_taxa", "pl_taxa")) |> 
    left_join(metaweb_long, by = c("bf_taxa", "pl_taxa")) |> 
    mutate(int_evidence = case_when(
        interaction == 1 ~ "recorded",
        int_potential == 1 ~ "potential",
        TRUE ~ "absence"
    ))
```

Load support matrix and join with the df:
```{r}
support_mat <- OmegaCor[[1]]$support

# keep only upper triangle of the matrix
support_mat[lower.tri(support_mat, diag = TRUE)] <- NA

support_long <- support_mat |>
    as.data.frame() |>
    rownames_to_column("bf_taxa") |>
    pivot_longer(cols = -bf_taxa, names_to = "pl_taxa", values_to = "support") |> 
    filter(!is.na(support)) |> 
    filter(bf_taxa %in% butterflies & pl_taxa %in% plants)

df <- df |> 
    left_join(support_long, by = c("bf_taxa", "pl_taxa"))
```

Overview:

```{r}
df |> group_by(interaction) |> 
    summarise(n = n(),
              mean_residuals = mean(residuals), 
              sd_residuals = sd(residuals),
              mean_support = mean(support),
              sd_support = sd(support)) |> 
    arrange(desc(mean_residuals))
# NA of interaction come from the genus Vulpia which is not reported in Clarke (neither in the potential df). Although Vulpia is supposed to be a synonim of Festuca which is indeed recorded in Clarke
```

Distribution plot of residuals for observed interactions:
```{r}
ggplot(df|> filter(!is.na(interaction)), 
        aes(x = residuals, y = 1, 
            fill = as.factor(interaction), 
            color = as.factor(interaction))) +
    ggdist::stat_halfeye(alpha  = 0.5) +
    theme_minimal() + theme(legend.position = "none") 
```

Distribution plot of residuals for observed AND POTENTIAL interactions:
```{r}
ggplot(df |> filter(!is.na(int_potential)), 
        aes(x = residuals, y = 1, 
            fill = as.factor(int_potential), 
            color = as.factor(int_potential))) +
    ggdist::stat_halfeye(alpha  = 0.5) +
    theme_minimal() + theme(legend.position = "none") 
```

Distribution plot of residuals SEPARATING observed and potential interactions:
```{r}
ggplot(df |> filter(!is.na(int_potential)), 
        aes(x = residuals, y = 1, 
            fill = int_evidence, 
            color = int_evidence)) +
    ggdist::stat_halfeye(alpha  = 0.5) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

Plot relation of residuals with support for the different level of interaction evidence:
```{r}
# hist(df$interaction)
# hist(df$int_potential)
# hist(df$residuals)
# hist(df$support)

ggplot(df|> filter(!is.na(int_potential)), 
           aes(x = support, y = residuals, 
               fill = as.factor(int_evidence), 
               color = as.factor(int_evidence))) +
    geom_point(alpha = 0.2) +
    facet_wrap(~int_evidence) +
    theme_minimal() + theme(legend.position = "none")
```

## Model for testing interaction effect on residuals
```{r}
#| message: false
#| warning: false
library(glmmTMB)

hist(df$residuals)

m1 <- glmmTMB(residuals ~ int_evidence + (1 | bf_taxa), family = gaussian, data = df)
summary(m1)
# performance::check_model(m1)
DHARMa::simulateResiduals(m1, plot = TRUE)
visreg::visreg(m1, "int_evidence", scale = "response", gg = TRUE) + theme_minimal()
emmeans::emmeans(m1, pairwise ~ int_evidence)
```

```{r}
m2 <- glmmTMB(residuals ~ int_evidence + (1 | bf_taxa) + (1 | pl_taxa), family = gaussian, data = df)
m3 <- glmmTMB(residuals ~ int_evidence, family = gaussian, data = df)
summary(m2)
DHARMa::simulateResiduals(m2, plot = TRUE)
```
